ANTS: Advanced Open-Source Tools for Normalization And Neuroanatomy

### 2009-06-11 Documentation/Tex/antstheory.tex
Moved things around.  Added text regarding experiments.  Added a couple figures.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@25 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
\author[label1]{PICSL}

% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
%\address[label2]{Department of Mathematics, University of Pennsylvania, Philadelphia, PA, USA}

@@

We first provide an overview of the different components included in ANTS, such as the available transformation models and similarity metrics offered.  We follow this up with an extensive validation \ldots

@@

\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeo.}
           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Transformations and a subset of the similarity metrics available in ANTS.  Similarity metric acronyms:  CC = cross correlation, PR = probabilistic matching, MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.  The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.  All 
ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
  \label{table:chart}
\end{table*}    
\vspace{0in}

@@

\begin{figure}
  \centering
    \begin{tabular}{c}
      % \includegraphics[width=80mm]{hierarchy}
    \end{tabular}
  \caption{Diagrammatic illustration of the transformation model hierarchy where the encompassing transformation spaces are characterized by increasing degrees of freedom.  }
  \label{fig:hierarchy}
\end{figure}

\subsubsection{Linear Transformations--Rigid and Affine:}
Image registration strategies often begin with a linear transformation
for initial global alignment which follows a deformable
transformation with increased degrees of freedom.  
Affine transformations within ANTS optimize either a
mean-squared difference metric or a mutual information (MI) image
similarity metric.  ANTS affine transformations are optimized
w.r.t. translation, rotation, scaling and shearing in succession.
This strategy allows careful control of each component of the affine
transformation.  Internally, ANTS also composes the affine
transformation with the deformation field before performing any
interpolation or downsampling.  In this way, ANTS normalization never
requires more than a single image interpolation step and is able to
always refer back to the original full-resolution image sources.

\subsubsection{Diffeomorphic Transformations}
In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse (\cite{Ebin1970,Mumford1998}).  Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.

ANTS assumes the diffeomorphism (differentiable mapping with a differentiable
inverse), $\phi$, is defined on the image domain, $\Omega$, and
maintains an affine transform at the boundary such that
$\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an
affine mapping applied to the identity transformation.  The $\phi$,
over time, parameterizes a family of diffeomorphisms,
$\phi(\mathbf{x}, t) : \Omega \times t
\rightarrow \Omega$, which can be generated by integrating a
time-dependent, regularized velocity field, $\boldsymbol{v} : \Omega \times
t \rightarrow \mathbb{R}^d$, through an ordinary differential equation
(o.d.e.).  
\begin{align} \label{eq:ode}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t), t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.  
The deformation field provided by $\phi$ is
$\mathbf{u}(\mathbf{x})=\phi(\mathbf{x}, 1) - \mathbf{x}$.

Dupuis et al. (\cite{Dupuis1998}) motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}(x,t)||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis (\cite{Dupuis1998}) also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).  Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  (\cite{Dupuis1998,Miller2002})
\begin{align} \label{eq:lddmm}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2  dt + \lambda || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  ||  \right\}.
\end{align}
The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}).   The similarity metric, or data term, for LDDMM is the {\em squared intensity difference} with weight $\lambda$.

In medical image registration, one typically encounters more complex intensity transfers 
between one anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.  Thus, 
ANTS enables a variety of similarity terms beyond the squared difference:
\begin{align} \label{eq:diff}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2 dt  +~\lambda~\Pi( \mathcal{I}, \phi(\mathbf{x},1) , \mathcal{J} )  \right\}.
\end{align}
Here, $\Pi$ is a similarity metric depending on the images and the
mapping and $\lambda$ controls the degree of exactness in the
matching.  If $\Pi$ is selected as cross-correlation, then one is
estimating the diffeomorphism under more robust illumination
constraints, as described in \cite{Avants2008}.  A similar equation
for elastic/FFD/DMFFD matching may be minimized in ANTS.

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, Avants et al. construct a {\em symmetric} alternative to Equation (\ref{eq:diff}) (\cite{Avants2007b}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
\begin{align}\label{eq:diffs}
  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  =
     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
      \Bigg\{ & \int_0^{0.5} ( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   )dt  \nonumber \\
     &+ || \mathcal{I} \circ \phi_1(\mathbf{x},0.5) - \mathcal{J}\circ \phi_2(\mathbf{x},0.5)   ||\Bigg\}.
\end{align}
The corresponding symmetric Euler-Lagrange equations are similar to (\cite{Miller2002}).  Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$.

ANTS allows a variety and combination of data terms to be optimized
with diffeomorphisms and, at the same time,
enables a variety of models for parameterizing the $\phi(\cdot)$.  
Briefly, ANTS Diff models are: 
\begin{enumerate}
\item {\bf Exponential Mapping:}  This constant velocity field parameterization does not produce a distance metric; maps are not dense in the space of diffeomorphisms and are not inverse consistent.
\item {\bf Greedy SyN:} Approximates the distance metric with a fast approach.  Evaluated in \cite{Klein2009}.  Enforces $\phiinv(\phi(\x,1))=\x$ in the discrete domain.  Theoretically and computationally inverse consistent / symmetric. 
\item {\bf Geodesic SyN:}  Symmetrically minimizes the velocity field / distance metric explicitly over all time.  Computationally more expensive than, but produces similar solutions to, Greedy SyN for most medical imaging problems.
\end{enumerate}
All of these transformation models may be used with any of the ANTS similarity metrics.  In particular, ANTS allows $\Pi$ 
to be composed of a summation/weighted combination of available similarity metrics/data terms.

\paragraph{Exponential Mapping}
The key difference between a time-varying diffeomorphism and a diffeomorphism 
generated by an exponential mapping (Ashburner200x) is that the exponential 
mapping maintains only a single vector field that is constant in time.   Theoretically, 
the restricting the velocity field to be constant in time reduces the size of the 
space that may be generated \cite{ArnoldODE} in a way that is similar to the difference 
between real and rational numbers, the latter of which are sparsely distributed through 
the reals.  
In contrast, 
the full diffeomorphic space may be explored by using time-varying velocity fields 
as does SyN and LDDMM.  The exponentiation of a constant velocity field generates a 
diffeomorphism through, 
\begin{align} \label{eq:odec}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x},t) ), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
Note that there is no time parameter in the velocity field, here.   The Dartel algorithm uses 
the well-known ``scaling and squaring'' approach to efficiently generate $\phi$ from a velocity field $\boldsymbol{v}(x)$.

\paragraph{Greedy SyN}
Greedy optimization of equation~\ref{eq:diffs}, involves gradient and update steps 
as follows, 
\begin{align}
\text{gradient:   }  \nabla E = \partial_{\phi_i} \Pi ( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) ) \nonumber \\ 
\text{regularize:  } \nabla E \leftarrow G_\sigma \star \nabla E \nonumber \\ 
\text{update:  }  \phi_i(\x,0.5) \leftarrow \phi_i(\x,0.5)+\delta \nabla E ( \phi_i(\x,0.5) ). \nonumber \\
\text{update the inverse s.t.:  } \phi_i(\phiinv_i) = \Id.  
\end{align}
Though the gradients differ for each metric, the update scheme is the
same.  In words, the gradient is taken at the mid-point of the full
diffeomorphism, $\phi$, by decomposing with respect to its components,
$\phi_1$ and $\phi_2$.  The gradient with respect to each half is
computed, regularized with a Gaussian of variance $\sigma$ and then
mapped back to the origin of each diffeomorphism to update the full
mapping.  Hermosillo provides a good discussion of the gradients of
the similarity metrics available in ANTS
\cite{hermosillo}.  Note that, here, we supply a specific 
regularization scheme but others are available in ANTS.  
That is, one may directly regularize the $\phi_i$ or use 
no regularization at all.

\paragraph{Time-Varying SyN}
Global in time optimization of equation~\ref{eq:diffs}, involves a global gradient 
where, at each time point $t \in [0, 1]$, we have:
\begin{align}
\text{gradient:   }  \nabla E(\x,t) = \partial_{\phi_i} \Pi ( \mathcal{I}(\phiinv_1(\x,t)) ,  \mathcal{J}(\phiinv_2(\x,1-t)) ), \nonumber \\ 
\text{update:  }  \velocity(\x,t)=\velocity(\x,t)+G_\sigma \star \nabla E(\x,t). \nonumber \\
\text{reintegrate the o.d.e.s, via Runge-Kutta},  \nonumber \\
\text{that generate the $\phi_i$ in $[0,1]$.}  
\end{align}
The time-varying SyN algorithm is theoretically cleaner than greedy
SyN.  However, the computational and memory cost is greater due to the
dense-in-time gradient and requisite reintegration of the
diffeomorphisms after each update.  Note that, here, we supply a
specific regularization scheme but others are available in ANTS.

%\input{vectorspace.tex}

\subsubsection{Vector Field Operators:} Deformable normalization
strategies typically regularize a deformation, $\displace$, or
velocity field, $\velocity$, with a linear operator such as the
Laplacian or Navier-Stokes operator. One may write this regularization
as a variational minimization in terms of its linear operator or in
terms of a kernel function operating of the field itself, e.g., $
\displace_{smooth}= K \star \displace_{not-smooth}$, where $K$ is 
the Green's kernel for the linear operator, $L$ and $\star$ is
convolution.  ANTS enables a variety of choices for the kernel, $K$,
including the Gaussian with varying $\sigma$ and a variety of BSpline
functions, both of which induce adequate regularity for normalization
models used in ANTS.  Regularization models operate on either/both the
whole mapping, $\phi$, or/and the gradient of the similarity term.
The same schema is applied for diffeomorphic and the recently proposed directly-manipulated 
free-form-deformation
(DMFFD)\cite{Tustison2009}, allowing regularization of both total 
deformation and deformation update.  While additional physical operators will be
included in future releases, current BSpline options provide many orders 
of flexibility \cite{Tustison2005}.

@@

A simple and efficient, yet powerful image normalization algorithm is Thirion's demons (\cite{Thirion1998}).  Using an optical flow based similarity, the solution is obtained by iterating between the calculation of image forces and subsequent Gaussian regularization.  In ANTS, we extend this basic approach to include both cross-correlation and mean-squared difference.

@@

For $n$-D images, the FFD transformational model, $\phi_{FFD}$, is defined as 
\begin{align}\label{eq:TFFD}
  \phi_{FFD} =  \sum_{i_1=1}^{M_1}\ldots\sum_{i_n=1}^{M_n} \mathbf{P}_{i_1, \ldots, i_n} \prod_{j=1}^nB_{i_j,d_j}(u_{j})     
\end{align}
where $\mathbf{P}_{i_1, \ldots, i_n}$ is an $n$-D grid of control points and $B_{i_j,d_j}(u_{j})$ is the B-spline in the $i_j^{th}$ direction of order $d_j$.

@@

\subsection{ANTS Intensity Based Similarity Metrics}
Several intensity-based image metrics have been proposed in the literature with varying levels of performance dependent upon specific applications.  We have included several of the most popular similarity metrics within ANTS.  In addition, our software framework facilitates the development of other image metrics if that is desired by the user.

>>>>

\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
%\author[label1]{Brian B. Avants}
%\author[label1]{Nicholas J. Tustison}
%\author[label1]{Gang Song}
%\author[label1]{James C. Gee}
\author{Brian B. Avants}
\author{Nicholas J. Tustison}
\author{Gang Song}
\author{James C. Gee}% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
%\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
\address{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
%\address[label2]{Department of Mathematics, University of Pennsylvania, Philadelphia, PA, USA}

@@

We first provide an overview of the different components included in ANTS, such as the available transformation models and similarity metrics offered.  This is followed by an extensive experimental analysis that builds upon the results from the recent image normalization evaluation of \cite{Klein2009} which was limited to a single configuration of ANTS.  This will provide  a useful benchmark for evaluating other possible ANTS configurations.

@@

\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & translation and rotation \\
       {} & Affine$^\dagger$ & MI, MSQ & rigid, scaling, and shear \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, JHCT, PSE & Demons-like algorithm \\
       {} & DMFFD & CC, PR, MI, MSQ, JHCT, PSE  & FFD variant \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeomorphic}
           & Exponential &  CC,PR, MI, MSQ, JHCT, PSE  & minimizes $\boldsymbol{v}(\mathbf{x})$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, JHCT, PSE  & minimizes $\boldsymbol{v}(\mathbf{x,t})$ locally in time \\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, JHCT, PSE  & minimizes $\boldsymbol{v}(\mathbf{x,t})$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Available transformations and similarity metrics available in ANTS.  Similarity metric acronyms:  CC = cross correlation, PR = probabilistic matching, MSQ = mean squared difference, MI = mutual information, JHCT = Jensen-Havrda-Charvat-Tsallis divergence, PSE = point-set expectation.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.}
  \label{table:chart}
\end{table*}

@@

\begin{figure}
  \centering
    \begin{tabular}{c}
      \includegraphics[width=80mm]{hierarchy}
    \end{tabular}
  \caption{Diagrammatic illustration of the transformation model hierarchy where the encompassing transformation spaces are characterized by increasing degrees of freedom.  }
  \label{fig:hierarchy}
\end{figure}

\subsubsection{Rigid and Affine Linear Transformations }

Image registration strategies often begin with a linear transformation for initial global alignment which precedes a deformable transformation with increased degrees of freedom.  The linear transformations available within ANTS optimize either a mean-squared difference (MSQ) or mutual information (MI) similarity metric which are optimized with respect to translation, rotation, and, in the case of affine transformations, scaling and shearing.  The successive optimization of each component of the linear transformation allows for careful control over increasing degrees of freedom.   ANTS also composes the affine transformation with the deformable transformation field before performing any interpolation or downsampling.  In this way, ANTS normalization never requires more than a single image interpolation step and is able to always refer back to the original full-resolution image sources.    
%We have implemented quaternion-based rigid and affine transformations within ANTS.  Not only does this facilitate accessibility of a sound global alignment strategy, but the parameters of the calculated optimal linear transformation can be maintained internally to ANTS to avoid unnecessary image interpolations.  Either a mean-squared difference metric or a mutual information metric can be used for the linear alignment.

\subsubsection{Vector Field Operators for Regularization} Antecedent to discussion of available deformable transformation models, we point out that  deformable normalization strategies typically invoke a deformation regularization step which smooths the displacement field, $\displace$, or velocity field, $\mathbf{v}$,  or both by a linear operator such as the Laplacian or Navier-Stokes operator. One may write this regularization as a variational minimization in terms of its linear operator or in
terms of a kernel function operating of the field itself, e.g., $
\displace_{smooth}= K \star \displace_{not \,\, smooth}$, where $K \star$ denotes 
convolution with the Green's kernel, $K$, for the linear operator, $L$.

Regularization models operate on either the
whole mapping $\phi$ or the gradient of the similarity term or both.
The same regularization schema is available for both diffeomorphic and the recently 
proposed directly manipulated free-form deformation
(DMFFD) \cite{Tustison2009}, allowing regularization of both total 
deformation and deformation update.  Viewed from this perspective, hybrid configurations incorporating discretized FFD strategies and diffeomorphisms can be combined for novel 
image normalization approaches.

ANTS enables a variety of choices for $K$
including the Gaussian with varying $\sigma$ and a variety of B-spline
functions, both of which induce adequate regularity for normalization
models used in ANTS.  While additional physical operators will be
included in future releases, current B-spline options provide many orders 
of flexibility \cite{Tustison2005}.

\subsubsection{Diffeomorphic Transformations}

In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse (\cite{Ebin1970,Mumford1998}).  Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.

ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, an maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.  $\phi$, over time, parameterizes a family of diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow  \Omega$, which can be generated by integrating a time-dependent, smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow \mathbb{R}^d$,  through the ordinary differential equation (o.d.e.)
\begin{align} \label{eq:ode}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t), t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.  The deformation field yielded by $\phi$ is $\mathbf{u}(\mathbf{x}) = \phi(\mathbf{x},1) - \mathbf{x}$.

Dupuis et al. (\cite{Dupuis1998}) motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis (\cite{Dupuis1998}) also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).  Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  (\cite{Dupuis1998,Miller2002})
\begin{align} \label{eq:lddmm}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2dt  + \lambda\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  || d\Omega \right\}.
\end{align}
The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}) with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.

To accommodate a variety of medical image normalization tasks, one typically encounters more complex intensity transfers between one anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.  Thus, 
ANTS enables not only a variety of similarity metric possibilities beyond the conventional squared difference
metric but it also permits any number of different similarity metrics for a particular image normalization task.  
This leads to the following generalization of  Equation (\ref{eq:lddmm}):
\begin{align} \label{eq:diff}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2 dt  +\lambda\int_{\Omega}\Pi_{\sim}( \mathcal{I}, \phi(\mathbf{x},1) , \mathcal{J} )d\Omega  \right\}
\end{align}
where $\Pi_{\sim}$ is a similarity metric depending on the images and the
mapping and $\lambda$ controls the degree of exactness in the
matching.  If $\Pi_{\sim}$ is selected as cross-correlation, then one is
estimating the diffeomorphism under more robust illumination
constraints, as described in \cite{Avants2008}.  
%A similar equation for elastic/FFD/DMFFD matching may be minimized in ANTS.

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, \cite{Avants2007b} construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
\begin{align}\label{eq:diffs}
  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  = &
     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
      \Bigg\{ \nonumber \\
      & \int_0^{0.5} \left( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   \right)dt  \nonumber \\
     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
\end{align}
The corresponding symmetric Euler-Lagrange equations are similar to (\cite{Miller2002}).  Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).

Other diffeomorphic flavors have since been reported in the research literature (e.g. DARTEL (\cite{Ashburner2007}) and Diffeomorphic Demons (\cite{Vercauteren2007,Vercauteren2009}).  We include three diffeomorphic transformation models for parameterizing $\phi(\cdot)$.  These include Geodesic SyN, Greedy SyN, and exponential mapping.  As summarized in Table \ref{table:chart}, each of these transformation models can utilize a host of similarity measures both individually and in mutual combination.

%Briefly, ANTS diffeomorphic transformation models are: 
%\begin{enumerate}
%\item {\bf Exponential Mapping:}  This constant velocity field parameterization does not produce a distance metric; maps are not dense in the space of diffeomorphisms and are not inverse consistent.
%\item {\bf Greedy SyN:} Approximates the distance metric with a fast approach.  Evaluated in \cite{Klein2009}.  Enforces $\phiinv(\phi(\x,1))=\x$ in the discrete domain.  Theoretically and computationally inverse consistent / symmetric. 
%\item {\bf Geodesic SyN:}  Symmetrically minimizes the velocity field / distance metric explicitly over all time.  Computationally more expensive than, but produces similar solutions to, Greedy SyN for most medical imaging problems.
%\end{enumerate}
%All of these transformation models may be used with any of the ANTS similarity metrics.  In particular, ANTS allows $\Pi$ 
%to be composed of a summation/weighted combination of available similarity metrics/data terms.

\paragraph{Geodesic SyN}
Using a gradient-based optimization strategy, minimization implied by Equation (\ref{eq:diffs}), which we denote as $\Pi$, first requires a specified discretization of $t \in [0,1]$.  Calculation of the gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and $\phi_2$ is performed at each of the $t$ values which is given by
\begin{align}
  \nabla \Pi(\boldsymbol{x},t) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t)) 
\end{align}
for $i \in \{1,2\}$.   The corresponding velocity fields at each $t$ are then updated from the previous iteration according to 
\begin{align}
  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + K \star \nabla \Pi(\mathbf{x},t).
\end{align}
We then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in \{1,2\}$.  We cycle through these steps until convergence or iterative exhaustion.

\paragraph{Greedy SyN}
Although the Geodesic SyN algorithm conforms most closely to the theoretical diffeomorphic foundations culminating with Equation (\ref{eq:diffs}), the computational and memory cost is significant due to the dense-in-time gradient calculations and requisite reintegration of the diffeomorphisms after each iterative update.  As a lower-cost alternative, we offer a greedy variant which performs quite well for most medical image normalization problems we have encountered.  Additionally, this was the strategy used in the large-scale comparative image registration algorithm assessment of \cite{Klein2009}.

Greedy optimization of Equation (\ref{eq:diffs}) requires calculation of the gradient of the gradient taken at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$
\begin{align}
\nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
\end{align}
for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and $\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration according to 
\begin{align}
  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+\delta  (K \star \nabla \Pi ( \phi_i(\x,0.5) )).
\end{align}
where $\delta$ is a user-specifed step parameter.  The gradient is then mapped back to the origin of each diffeomorphism to update the full mapping by enforcing $\phiinv(\phi(\x,1))=\x$.

\paragraph{Exponential Mapping}
Ashburner introduced DARTEL (Diffeomorphic Anatomical Registration using Exponentiated Lie algebra) as a rapidly computed alternative to time parameterized diffeomorphic schemes (\cite{Ashburner2007}).  The key difference between a time-varying diffeomorphism and a diffeomorphism generated by an exponential mapping (\cite{Ashburner2007}) is that the exponential mapping maintains only a single vector field that is constant in time.

By exponentiation of a constant velocity field, a diffeomorphism can be generated from the following o.d.e (cf Equation (\ref{eq:ode}))
\begin{align} \label{eq:odec}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x},t) ), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
Note that there is no explicit time parameter in the velocity field.   Theoretically, restricting the velocity field to be constant in time reduces the size of the space that may be generated (\cite{Arnold1991}) in a way that is similar to the difference between real and rational numbers, the latter of which are sparsely distributed through the reals.

@@

A simple and efficient, yet powerful image normalization algorithm is the approach known as Thirion's demons (\cite{Thirion1998}).  Using an optical flow based similarity, the solution is obtained by iterating between the calculation of image forces and subsequent Gaussian regularization.  In ANTS, we extend this basic approach to include the similarity metrics available for deformable registration (see Table \ref{table:chart}).

@@

For $n$-D images, the FFD (and DMFFD) transformational model, $\phi_{FFD}$, is defined as 
\begin{align}\label{eq:TFFD}
  \phi_{FFD} =  \sum_{i_1=1}^{M_1}\ldots\sum_{i_n=1}^{M_n} \mathbf{P}_{i_1, \ldots, i_n} \prod_{j=1}^nB_{i_j,d_j}(u_{j})     
\end{align}
where $\mathbf{P}_{i_1, \ldots, i_n}$ is an $n$-D grid of control points and $B_{i_j,d_j}(u_{j})$ is the B-spline in the $i_j^{th}$ direction of order $d_j$.

@@

\subsection{ANTS Intensity-Based Similarity Metrics}
Several intensity-based image metrics have been proposed in the literature with varying levels of performance dependent upon specific applications.  We have included several of the most popular similarity metrics within ANTS.  In addition, our software framework facilitates the development of other image metrics.
